### 2025-08-28 manuscript/manuscript.qmd
updated manuscript (abstract) started discussion. added more to keyword analysis

abstract: "This study examines the gender distribution of authors in music psychology. A total of 3,373 papers published between 2000 and 2025 across five journals were analyzed with respect to author gender and authorship role (first, last, solo, coauthor). In addition, patterns related to citations, open-access publishing, and keywords were explored across gender. Overall, female authors were in the minority, accounting for 40.2% of all authors. However, the distribution of authorship roles diverged notably from this overall proportion: women were more likely to be first authors (OR = 1.41), equally likely to be solo authors (OR = 1.11) or coauthors (OR = 1.00), but less likely to be last authors (OR = 0.73). These patterns have become more pronounced over the past 15 years. No substantial gender differences in citation counts were observed. Analyses of geographical variation revealed clear cross-country differences, potentially reflecting both the status of women and the historical development of the discipline in particular regions. Finally, an examination of keywords suggested broad thematic preferences differing by gender."  
keywords          : "gender, transparency, music psychology, open science, meta-research"
wordcount         : "5555"
figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : no
mask              : no
csl: apa7.csl
class             : "man"
bibliography: references.bib
format:
  html: default
---

@@

Research across several scientific disciplines that show that women tend to be less represented in authors compared to men [@banks2025women; @son2022scientific]. Encouragingly, however, there is a general trend that the ratio of women to men in authorships is improving. From around 1960 to 2021, studies show a general improvement [@gonzalez2020women; @ioannidis2023gender; @jemielniak2025gender]. However, there is evidence that this improvement is plateauing [@jemielniak2025gender] and that although women are more likely to be first author, women are still generally unfairly underrepresented as the last author [@gonzalez2020women; @rock2021quantifying; @shah2021gender]. There could be several reasons for such a disparity: despite efforts to give fair credit for author contributions [@ni2021gendered], for example, using the Contributor Roles Taxonomy (CRediT) system, assigning authorships can be unclear. What counts as an authorship may also differ between academic disciplines. Women are more likely to experience authorship disagreements [@ni2021gendered]. There are general disagreements of when to decide authorship - women preferring to discuss in earlier stages, but men choosing authorship at the final stage. Another reason could also come from parenthood and parental leave, which account for about ~40% of gender gap in career advancement [@nielsen2024getting]. Importantly, there are strides to improving this, with opinion papers discussing way to improve receipt and reporting of intellectual credit [@banks2025women] as well as creating equitable environments in academic science [@martinez2024comparison].

@@

We carried out manual corrections of the gender attributions where we were familiar with the author or the database had the full first name coded with initials, or the second name used as the first name. This led to `32` corrections.

@@

For the analyses, we analysed gender disparities in authorship by comparing the frequency of these authorship types between male and female authors. Our analysis focused on the female author proportion (FAP) and odds ratios comparing the likelihood of occupying each authorship position by gender.

@@

The overall odds ratio for women as first authors was 1.41 [95% CI: 1.28–1.54], indicating that the odds of first authorship are 41% higher for females compared to the male authors (Fisher's exact test _p_ < .001). For coauthorship positions, the OR was 1.00 [0.91–1.09], suggesting no substantial gender difference (_p_ = .96). In contrast, the odds ratio for women as last authors was 0.73 [0.67–0.81], indicating that the odds of last authorship are 27% lower for women compared to men and significantly different (_p_ < .001). Being a single author has odds ratio of 0.90 [0.78-1.05] for females, which suggests that while women are represented with relatively more single authored papers than men, the difference is not significant (_p_ = .21).

@@

To explore the potential differences within music psychology topics of choice between female and males, we analysed the keywords in the articles (15,316 in total, 6730 unique). To reduce variant keywords, we manually simplified the variants (e.g., "arts in health", "healthy music use", "health musicking" were converted to "music health" and "music cognition" to "cognition" and "music perception" to "perception") and eliminated single use of the term "music". We aggregated the counts of keywords across all authors in the articles and Figure 4 shows the odds ratios for female authors to be associated (in any author position) with the 40 most frequently used keywords. The keywords signifying related to development, emotion regulation, well-being, music performance anxiety, motivation, expertise, music listening, mental health, and music therapy show significantly higher odds ratios (OR $\ge$ 1.50) for females in comparison to overall gender ratios in the data. At the opposite end of this continuum, topics such as music analysis, machine learning, harmony, cognition, rhythm, and performance expression show the opposite trend, suggesting that the articles with these keywords tend to have overall more prevalent male authorships (female OR $\le$ 0.75).

@@

>>>>

abstract: "Academia suffers from issues regarding equity, with one  crucial aspect being academic authorship. Many studies have shown that there is a gender gap in publishing, where women are often underrepresented. As there are differences across disciplines, the goal of this study is to provide a current overview the gender distribution of authors within music psychology. A total of 3,373 papers published between 2000 and 2025 across five journals were analyzed with respect to author gender and authorship role (first, last, solo, coauthor). In addition, patterns related to citations, open-access publishing, and keywords were explored across gender. Overall, female authors were in the minority, accounting for 40.2% of all authors. However, the distribution of authorship roles diverged notably from this overall proportion: women were more likely to be first authors (OR = 1.41), equally likely to be solo authors (odds ratio, OR = 1.11) or coauthors (OR = 1.00), but less likely to be last authors (OR = 0.73). These patterns have become more pronounced over the past 15 years. No substantial gender differences in citation counts were observed. Analyses of geographical variation revealed clear cross-country differences, potentially reflecting both the status of women and the historical development of the discipline in particular regions. Finally, an examination of keywords suggested broad thematic preferences differing by gender. While overall some of this work shows that the field of music psychology is more balanced than some other field, especially in more STEM disciplines, providing such a picture hopefully inspires discussions and initiatives to promote fairness and more equity in academia."  
keywords          : "gender, transparency, music psychology, open science, meta-research"
wordcount         : "5555"
figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : no
mask              : no
csl: apa7.csl
class             : "man"
bibliography: references.bib
format:
  html: default
---

@@

Research across several scientific disciplines that show that women tend to be less represented in authors compared to men [@banks2025women; @son2022scientific]. Encouragingly, however, there is a general trend that the ratio of women to men in authorships is improving. From around 1960 to 2021, studies show a general improvement [@gonzalez2020women; @ioannidis2023gender; @jemielniak2025gender]. Yet there is evidence that this improvement is plateauing [@jemielniak2025gender] and that although women are more likely to be first author, women are still generally unfairly underrepresented as the last author [@gonzalez2020women; @rock2021quantifying; @shah2021gender]. There could be several reasons for such a disparity: despite efforts to give fair credit for author contributions [@ni2021gendered], for example, using the Contributor Roles Taxonomy (CRediT) system, assigning authorship can be unclear. What counts as an authorship may also differ between academic disciplines. Women are more likely to experience authorship disagreements, for example, when to decide authorship - women prefer to discuss in earlier stages, while men choose authorship at the final stage [@ni2021gendered]. Another reason for the gender gap could also come from parenthood and parental leave, which account for about ~40% of gender gap in career advancement [@nielsen2024getting]. Importantly, there are strides to improving this, with opinion papers discussing way to improve receipt and reporting of intellectual credit [@banks2025women] as well as creating equitable environments in academic science [@martinez2024comparison].

@@

We carried out manual corrections of the gender attributions where we were familiar with the author or the database had the full first name coded with initials, or the second name used as the first name. This led to `53` corrections.

@@

For the analyses, we analysed gender disparities in authorship by comparing the frequency of these authorship types between male and female authors. Our analysis focused on the female author proportion (FAP) and odds ratios (ORs) comparing the likelihood of occupying each authorship position by gender.

@@

The overall OR for women as first authors was 1.41 [95% CI: 1.28–1.54], indicating that the odds of first authorship are 41% higher for females compared to the male authors (Fisher's exact test _p_ < .001). For coauthorship positions, the OR was 1.00 [0.91–1.09], suggesting no substantial gender difference (_p_ = .96). In contrast, the odds ratio for women as last authors was 0.73 [0.67–0.81], indicating that the odds of last authorship are 27% lower for women compared to men and significantly different (_p_ < .001). Being a single author has odds ratio of 0.90 [0.78-1.05] for females, which suggests that while women are represented with relatively more single authored papers than men, the difference is not significant (_p_ = .21).

@@

To explore the potential differences within music psychology topics of choice between female and males, we analysed the keywords in the articles (15,316 in total, 6730 unique). To reduce variant keywords, we manually simplified the variants (e.g., "arts in health", "healthy music use", "health musicking" were converted to "music health" and "music cognition" to "cognition" and "music perception" to "perception") and eliminated single use of the term "music". We aggregated the counts of keywords across all authors in the articles and Figure 4 shows the odds ratios for female authors to be associated (in any author position) with the 40 most frequently used keywords. The keywords signifying related to children, development, emotion regulation, well-being, music performance anxiety, motivation, expertise, music listening, mental health, and music therapy show significantly higher odds ratios (OR $\ge$ 1.50) for females in comparison to overall gender ratios in the data. At the opposite end of this continuum, topics such as music analysis, machine learning, harmony, cognition, rhythm, and performance expression show the opposite trend, suggesting that the articles with these keywords tend to have overall more prevalent male authorships (female OR $\le$ 0.75).

@@

# Discussion
Authorship is a crucial aspect in academia as a marker of scholarly success and strongly influences academic career progression, influencing recognition, funding, and advancement opportunities. However, there are several unequities within this aspect of academia, with women often being underrepresented. As such representations differ across different disciplines, it is worthwhile understanding how The goal of the current paper was to provide a snapshot of authorship across gender. While we recognise that gender is a complex and multidimensional social construct, and there are limitations to treating gender as a binary category, we used a binary male-female classification consistent with prior meta-science research on gender and authorship [@son2022scientific; @shah2021gender; @rock2021quantifying; @ni2021gendered; @gonzalez2020women; @wais2006]. Overall, we see a relatively encouraging pattern within
